---
title: "Taking Home Charlie"
author: "Robert Nguyen"
date: "7 September 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fitzRoy)
library(tidyverse)
library(ordinal)
library(MASS)
library(janitor)
```

## Build your own Brownlow Model

G'day all, your friendly neighbourhood PhD student here. I am here to help you all build your own models. 

That is very different from what you might be used to which could consist of any one of the following scenarios. 

* Here is a model I have build that you are not smart enough too, follow my tips blindly
* Here is a model that I have build with data you don't have access to, therefore it is better than what you can do follow my tips
* Here is a model that I don't explain well apart from performance, follow my tips

Hopefully you see my point and see the unfortunate pattern. 

This is **NOT** going to be one of those things, this is a guide to help you build a competitive Brownlow prediction model. It is fully reproducible and you can use the variables created if you want, you can create your own variables but by the end of this process the model is yours. 

So let me try to ease some possible thoughts.

1. Where can I get the data from? 

[James Day](https://twitter.com/jamesday87) and [myself](https://twitter.com/anoafl) have worked together to produce an R package called [fitzRoy](https://github.com/jimmyday12/fitzRoy) which means that you and anyone else can have easy access to all the good stuff you see on [afltables](https://afltables.com/afl/afl_index.html) and [footywire](https://www.footywire.com/). 

2. But what is the point other people do models why build my own when I can just use someone else's? 

That is a good point, there are Brownlow Models floating around, and you could even go to a tipping service and pay for tips. But I assume you are interested in predicting the Brownlow Medallist  because you are a football nuffie and as such have ideas about what *you* think have an effect on how umpires vote. 

3. So if I have the data and I can build my own model, but I don't want to because it won't be any good.

All models are wrong but some are useful -- This model is by no means perfect, that isn't the point. But it will provide you a predicted order that has less ranking error vs Champion Data's official prediction using data that you don't have access to without $$$$

4. So what?

In a nutshell, below is a fully reproducible method you can edit and tinker with that without any tinkering with produced a better top 10 Brownlow Medal prediction vs the official data provider without any of the bells and whistles!

## What Packages do I need?


```{r }
library(MASS)
library(ordinal)
library(fitzRoy)
library(tidyverse)
```

So now you have the packages you need, first steps lets get the data ready

## Getting the data 

First we can use the [fitzRoy](https://github.com/jimmyday12/fitzRoy) package to get all the data from [afltables](https://afltables.com/afl/seas/2018.html)

This will consist of basically 3 steps

1. Get the data from afltables, we can do this using the `fitzRoy::get_afltables_stats`

2. Create the game by game summaries so we have the total kicks per game, total handballs per game by team for all the statistics available on [afltables](https://afltables.com/afl/stats/games/2018/031420180322.html)

3. Join on scores - because people feel as though margin has an effect on the order of 3,2,1 (Hard to get the 3 votes in a flogging)

We do this using the `fitzRoy::match_results` and then we join the datasets together creating a final data frame that has all player statistics at an individual level, joined with their team totals and game score. 



```{r}
df<-fitzRoy::get_afltables_stats(start_date = "1897-01-01", end_date = Sys.Date())
df<-df%>%filter(Season>2010)

names(df)
team_stats<-df%>%
  dplyr::select(Date, First.name,Surname,Season, Round, Playing.for, Kicks:Goal.Assists)%>%
  group_by(Date, Season, Round, Playing.for)%>%
  summarise_if(is.numeric,funs(sum=c(sum(.))))

player_stats<-df%>%
  dplyr::select(Date, First.name,Surname,Season, Round, Playing.for, Kicks:Goal.Assists)

complete_df<-left_join(player_stats,team_stats, by=c("Date"="Date", "Season"="Season",  "Playing.for"="Playing.for"))

dataset_scores<-fitzRoy::match_results
names(dataset_scores)
dataset_scores1<-dataset_scores%>%dplyr::select (Date, Round, Home.Team, Home.Points,Game)
dataset_scores2<-dplyr::select(dataset_scores, Date, Round, Away.Team, Away.Points,Game)
#Sometimes when joining datasets together it helps to rename things for consistency

colnames(dataset_scores1)[3]<-"Team"
colnames(dataset_scores1)[4]<-"Points"
colnames(dataset_scores2)[3]<-"Team"
colnames(dataset_scores2)[4]<-"Points"

df5<-rbind(dataset_scores1,dataset_scores2)
dataset_margins<-df5%>%group_by(Game)%>%
  arrange(Game)%>%
  mutate(margin=c(-diff(Points),diff(Points)))
# View(dataset_margins) # I have commented this out, but always good to view 
dataset_margins$Date<-as.Date(dataset_margins$Date)
complete_df$Date<-as.Date(complete_df$Date)

complete_df<-left_join(complete_df,dataset_margins,by=c("Date"="Date",  "Playing.for"="Team"))

```

## Create the Ratios

Create ratios is important for the model, remember we are interested in their relative performance to their peers. Is it impressive kicking 3 goals if 25 have been kicked? Or is it more impressive getting 3 if only 5 goals have been kicked? 

Intuition  like this for kicks, handballs, clearances etc would lead us to believe that ratios are better predictors of polling votes over raw statistics. 

```{r}
complete_df_ratio<-complete_df%>%
  mutate(kick.ratio=Kicks/Kicks_sum,
         Marks.ratio=Marks/Marks_sum,
         handball.ratio=Handballs/Handballs_sum,
         Goals.ratio=Goals/Goals_sum,
         behinds.ratio=Behinds/Behinds_sum,
         hitouts.ratio=Hit.Outs/Hit.Outs_sum,
         tackles.ratio=Tackles/Tackles_sum,
         rebounds.ratio=Rebounds/Rebounds_sum,
         inside50s.ratio=Inside.50s/Inside.50s_sum,
         clearances.ratio=Clearances/Clearances_sum,
         clangers.ratio=Clangers/Clangers_sum,
         freefors.ratio=Frees.For/Frees.For_sum,
         freesagainst.ratio=Frees.Against/Frees.Against_sum,
         Contested.Possessions.ratio=Contested.Possessions/Contested.Possessions_sum,
         Uncontested.Possessions.ratio=Uncontested.Possessions/Uncontested.Possessions_sum,
         contested.marks.ratio=Contested.Marks/Contested.Marks_sum,
         marksinside50.ratio=Marks.Inside.50/Marks.Inside.50_sum,
         one.percenters.ratio=One.Percenters/One.Percenters_sum,
         bounces.ratio=Bounces/Bounces_sum,
         goal.assists.ratio=Goal.Assists/Goal.Assists_sum,
         disposals.ratio=(Kicks+Handballs)/(Kicks_sum+Handballs_sum))
df<-complete_df_ratio%>%dplyr::select(Date, First.name, Surname, Season, Round.x, Playing.for,-Brownlow.Votes, Brownlow.Votes_sum,everything())
df<-df%>%dplyr::select(-Brownlow.Votes,everything())
df[is.na(df)] <- 0 # we have to replace NA with 0


```


Now we have all that we need for the dataset now lets do some modelling!

## Fit a proportional odds model

Basically all we are doing is using the dataset we have and fitting a proportional odds model. What this means is that we will end up with a probability that each player will poll 3 votes, 2 votes, 1 vote and 0 votes for each and every single game. 

This will involve a few steps

1. use `subset` to get our training data
2. use `filter` to make sure we only have home and away games --this is a data knowledge step, we know that you can not get votes in finals, but the column instead of being left empty is filled with 0s so an easy way to make sure we only have home and away games is using Round as a filter.
3. standardise our variables using `scale`

```{r}
in.sample  <- subset(df, Season %in% c(2013:2016)) # our training data

in.sample$Brownlow.Votes <- factor(in.sample$Brownlow.Votes)

in.sample<-in.sample%>%filter(Round.x %in% c("1","2","3","4","5","6","7","8",
                                             "9","10","11","12","13","14","15","16","17","18","19","20","21","22","23","24"))


names(in.sample) 

in.sample$Player<-paste(in.sample$First.name,in.sample$Surname)

in.sample<-in.sample%>%dplyr::select(Player, Date, Season, Round.x, Playing.for, margin:Brownlow.Votes)
in.sample<-in.sample[-c(1,2)] #removed for 'cleaner look'

temp1<-scale(in.sample[,6:27])
in.sample[,6:27]<-temp1
#attributes(temp1)
temp1.center<-attr(temp1,"scaled:center")
temp1.scale<-attr(temp1,"scaled:scale")







fm1<-clm(Brownlow.Votes~ handball.ratio +  Marks.ratio +  
           disposals.ratio+  hitouts.ratio+
           freefors.ratio +  freesagainst.ratio +  tackles.ratio +	Goals.ratio +	behinds.ratio +	Contested.Possessions.ratio+
           Uncontested.Possessions.ratio +	clangers.ratio +	contested.marks.ratio +	marksinside50.ratio +
           clearances.ratio +	rebounds.ratio +	inside50s.ratio +	one.percenters.ratio +	bounces.ratio+
           goal.assists.ratio  +margin, 
         data = in.sample)

```

## Variable selection

Nothing fancy here, remember the point is that you can pick and choose what variables you want as your predictors, here I am just going to use stepwise selection and go backwards. 

But by all means, I am not saying this is the best way, I'm just getting to a result as quickly as I can with little thinking about the variables. 

Ideally you might think about the variables and test them out for fit but moving on.

```{r}
fm2<- stepAIC(fm1, direction='backward',type=AIC)

```



## Lets do the prediction

Instead of just taking the data from [afltables](https://afltables.com/afl/seas/2018.html) in 2017, lets do something just a little bit different. Lets use [footywire](https://www.footywire.com/) as our out.sample 

This might seem like a bit of a weird step, but it is really valuable. Footywire has more data than afltables for games, it contains stats like tackles inside 50 as an example. Also as an example within fitzRoy we essentially have 2 data sources for you to use. Either afltables or footywire. There are benefits to both for example afltables data goes further back, but doesn't have extra columns that footywire has that you might want to use for example supercoach scores. 



```{r}
names(fitzRoy::player_stats)
df_2017<-fitzRoy::player_stats%>%
  filter(Season==2017)

team_stats_out<-df_2017%>%
  dplyr::select(Date, Player,Season, Round, Team, CP:T5)%>%
  group_by(Date,Season, Round, Team)%>%
  summarise_if(is.numeric,funs(sum=c(sum(.))))

player_stats_out<-df_2017%>%
  dplyr::select(Date, Player,Season, Round, Team, CP:T5)


complete_df_out<-left_join(player_stats_out,team_stats_out, by=c("Date"="Date", "Season"="Season",  "Team"="Team"))



dataset_scores<-fitzRoy::match_results
names(dataset_scores)
dataset_scores1<-dataset_scores%>%dplyr::select (Date, Round, Home.Team, Home.Points,Game)
dataset_scores2<-dplyr::select(dataset_scores, Date, Round, Away.Team, Away.Points,Game)

colnames(dataset_scores1)[3]<-"Team"
colnames(dataset_scores1)[4]<-"Points"
colnames(dataset_scores2)[3]<-"Team"
colnames(dataset_scores2)[4]<-"Points"


df5<-rbind(dataset_scores1,dataset_scores2)
dataset_margins<-df5%>%group_by(Game)%>%
  arrange(Game)%>%
  mutate(margin=c(-diff(Points),diff(Points)))
dataset_margins$Date<-as.Date(dataset_margins$Date)
complete_df_out$Date<-as.Date(complete_df_out$Date)

dataset_margins<-dataset_margins %>%mutate(Team = str_replace(Team, "Brisbane Lions", "Brisbane"))

dataset_margins<-dataset_margins %>%mutate(Team = str_replace(Team, "Footscray", "Western Bulldogs"))


complete_df_out<-left_join(complete_df_out,dataset_margins,by=c("Date"="Date",  "Team"="Team"))

names(complete_df_out)

####create the new ratios
complete_df_ratio_out<-complete_df_out%>%
  mutate(kick.ratio=K/K_sum,
         Marks.ratio=M/M_sum,
         handball.ratio=HB/HB_sum,
         Goals.ratio=G/G_sum,
         behinds.ratio=B/B_sum,
         hitouts.ratio=HO/HO_sum,
         tackles.ratio=T/T_sum,
         rebounds.ratio=R50/R50_sum,
         inside50s.ratio=I50/I50_sum,
         clearances.ratio=(CCL+SCL)/(CCL_sum+SCL_sum),
         clangers.ratio=CL/CL_sum,
         freefors.ratio=FF/FF_sum,
         freesagainst.ratio=FA/FA_sum,
         Contested.Possessions.ratio=CP/CP_sum,
         Uncontested.Possessions.ratio=UP/UP_sum,
         contested.marks.ratio=CM/CM_sum,
         marksinside50.ratio=MI5/MI5_sum,
         one.percenters.ratio=One.Percenters/One.Percenters_sum,
         bounces.ratio=BO/BO_sum,
         goal.assists.ratio=GA/GA_sum,
         disposals.ratio=D/D_sum)




conforming<-complete_df_ratio_out%>%
  dplyr::select(Player, Date, Season, Round.x, Team, margin, 
                kick.ratio:disposals.ratio)

conforming$Brownlow.Votes<-0
out.sample=conforming

newdata   <- out.sample[ , -ncol(out.sample)]

newdata[,6:27]<-scale(newdata[,6:27],center=temp1.center,scale=temp1.scale) 

pre.dict    <- predict(fm2,newdata=newdata, type='prob')
pre.dict.m  <- data.frame(matrix(unlist(pre.dict), nrow= nrow(newdata)))
colnames(pre.dict.m) <- c("vote.0", "vote.1", "vote.2", "vote.3")

newdata.pred  <- cbind.data.frame(newdata, pre.dict.m)

```

## Expected Votes

A proportional odds model makes the players total probabilities  for each voting category sum to 1 for every player that the player plays in. 

That is for Dustin Martin in round 1 2017, $1= \sum_{k=0}^3 p_{k}$ which makes sense.

But we know that there is another harder  restriction. That is only one player can get 3 votes, only one player can get 2 votes and only one player can get the 1 vote. 

Now lets add in our much tougher restriction. This will mean the probability of polling 3 votes will sum to 1 across all players. We will call this the std (standardised) votes.




```{r}
#### Step 1: Get expected value on Votes
newdata.pred$expected.votes <- newdata.pred$vote.1 + 2*newdata.pred$vote.2 + 3*newdata.pred$vote.3

####Join back on matchID whoops! 

get_match_ID<-fitzRoy::player_stats

xx<-get_match_ID%>%dplyr::select(Date, Player, Match_id)
newdata.pred<-left_join(newdata.pred, xx, by=c("Date"="Date",  "Player"="Player"))



newdata.pred<-filter(newdata.pred, Date<"2017-09-01")


sum1 <- aggregate(vote.1~Match_id, data = newdata.pred, FUN = sum ); names(sum1) <- c("Match_id", "sum.vote.1");
sum2 <- aggregate(vote.2~Match_id, data = newdata.pred, FUN = sum ); names(sum2) <- c("Match_id", "sum.vote.2");
sum3 <- aggregate(vote.3~Match_id, data = newdata.pred, FUN = sum ); names(sum3) <- c("Match_id", "sum.vote.3");

#### Step 3: Add sum of each vote by matchId to big table
newdata.pred <- merge(newdata.pred, sum1, by = "Match_id")
newdata.pred <- merge(newdata.pred, sum2, by = "Match_id")
newdata.pred <- merge(newdata.pred, sum3, by = "Match_id")

#### Step 4: Add std1/2/3
newdata.pred$std.1  <- (newdata.pred$sum.vote.1/newdata.pred$vote.1)^-1
newdata.pred$std.2  <- (newdata.pred$sum.vote.2/newdata.pred$vote.2)^-1
newdata.pred$std.3  <- (newdata.pred$sum.vote.3/newdata.pred$vote.3)^-1


#### Step 5: Expected standard game vote
newdata.pred$exp_std_game_vote <- newdata.pred$std.1 + 2*newdata.pred$std.2 + 3*newdata.pred$std.3  


#### Step 6: List of winners

newdata.pred$PlayerName<-paste(newdata.pred$Player," ",newdata.pred$Team)
winners.stdgame   <- aggregate(exp_std_game_vote~PlayerName, data = newdata.pred, FUN = sum );
winners.stdgame   <- winners.stdgame[order(-winners.stdgame$exp_std_game_vote), ]
winners.stdgame[1:10, ]


```

## But how good is this prediction?

We want our predicted order to be as close to correct as possible, especially for the top 10. 

So lets use as a measure of accuracy ranking error which I will define as difference between actual place vs predicted place. For example, we have predicted Dustin Martin to finish first and he did so error is 0, if we predicted Dustin Martin to be runner up (like Champion Data did) then we would have a ranking error of 1. 


```{r}

library(tidyverse)
data.frame(stringsAsFactors=FALSE,
           Predictor = c("anoafl", "anoafl", "anoafl", "anoafl", "anoafl", "anoafl", "anoafl", "anoafl",
                         "anoafl", "anoafl", "anoafl", "anoafl", "anoafl", "anoafl", "CD", "CD", "CD", "CD", "CD",
                         "CD", "CD", "CD", "CD", "CD", "CD", "CD", "CD", "CD"),
           Player = c("Dustin Martin", "Patrick DangerField", "Tom Mitchell",
                      "Dayne Zorko", "Josh Kelly", "Rory Laird", "Zach Merrett", "Rory Sloan",
                      "Lance Franklin", "Nat Fyfe", "Trent Cotchin", "Clayton Oliver",
                      "Callan Ward", "Marcus Bontempelli", "Patrick DangerField", "Dustin Martin",
                      "Clayton Oliver", "Tom Mitchell", "Rory Sloan", "Callan Ward", "Zach Merrett",
                      "Marcus Bontempelli", "Josh Kelly", "Trent Cotchin", "Rory Laird", "Lance Franklin",
                      "Nat Fyfe", "Dayne Zorko"),
           Ranking.Error = c(0L, 0L, 0L, 12L, 1L, 38L, 5L, 1L, 4L, 2L, 11L, 28L, 8L,
                             5L, 1L, 1L, 23L, 1L, 2L, 22L, 5L, 0L, 3L, 41L, 30L, 6L, 4L, 10L)
)%>%ggplot(aes(x=Ranking.Error, y=Player))+geom_point(aes(colour=Predictor), size=5)+ theme(axis.text.y=element_text(size=12)) +ylab("") 


```

So what we can see here is that by taking just the basic data and making them into game ratios and thinking a bit more about probability and with little thought into variable selection (backwards) we can come up with a prediction that does just as well (if not a little better) as champion data (although obviously biased).


So hopefully now you get the idea, this isn't meant to be the very best method, but hopefully it does encourage you to instead of using just backwards aic to select your variables you can do something else instead. 

## What changes to make?

Probably the most obvious one is the variable selection step. 

Now you could simply enter in the variables you thought might have an impact on polling brownlow votes and to be honest this is a good way to do things. 
As a football fan you would hope that your domain knowledge does improve things. 

So how to do that?

Lets say you think that only contested possessions, inside 50s and clearances matter after all it is a midfielders award.

```{r, eval=FALSE}

fm1<-clm(Brownlow.Votes~	Contested.Possessions.ratio+
           
           clearances.ratio +	inside50s.ratio,
         data = in.sample)

```

Then you can run everything through as normal and see how your model goes at predicting. 

## Why use AIC, why not BIC?

That's a good point and exactly the point I would like to make lots of these choices can be changed and depends on use. The whole point if this exercise  was to show that relatively quickly and with only freely available  data you can do well in comparison  to say CD neural  net. 

But if you favour interpretable models, you might want to use BIC, it penalises model complexity more so you will end up with models that have less variables. 

That step is also quick to change

```{r, eval=FALSE}
fm2<- stepAIC(fm1, direction='backward',type=BIC)
```


## Lets do 2018

```{r}
library(MASS)
library(ordinal)
library(fitzRoy)
library(tidyverse)

df<-fitzRoy::get_afltables_stats(start_date = "1897-01-01", end_date = Sys.Date())
df<-df%>%filter(Season>2010)

names(df)
team_stats<-df%>%
  dplyr::select(Date, First.name,Surname,Season, Round, Playing.for, Kicks:Goal.Assists)%>%
  group_by(Date, Season, Round, Playing.for)%>%
  summarise_if(is.numeric,funs(sum=c(sum(.))))

player_stats<-df%>%
  dplyr::select(Date, First.name,Surname,Season, Round, Playing.for, Kicks:Goal.Assists)

complete_df<-left_join(player_stats,team_stats, by=c("Date"="Date", "Season"="Season",  "Playing.for"="Playing.for"))

dataset_scores<-fitzRoy::match_results
names(dataset_scores)
dataset_scores1<-dataset_scores%>%dplyr::select (Date, Round, Home.Team, Home.Points,Game)
dataset_scores2<-dplyr::select(dataset_scores, Date, Round, Away.Team, Away.Points,Game)

colnames(dataset_scores1)[3]<-"Team"
colnames(dataset_scores1)[4]<-"Points"
colnames(dataset_scores2)[3]<-"Team"
colnames(dataset_scores2)[4]<-"Points"

df5<-rbind(dataset_scores1,dataset_scores2)
dataset_margins<-df5%>%group_by(Game)%>%
  arrange(Game)%>%
  mutate(margin=c(-diff(Points),diff(Points)))
# View(dataset_margins)
dataset_margins$Date<-as.Date(dataset_margins$Date)
complete_df$Date<-as.Date(complete_df$Date)

complete_df<-left_join(complete_df,dataset_margins,by=c("Date"="Date",  "Playing.for"="Team"))


complete_df_ratio<-complete_df%>%
  mutate(kick.ratio=Kicks/Kicks_sum,
         Marks.ratio=Marks/Marks_sum,
         handball.ratio=Handballs/Handballs_sum,
         Goals.ratio=Goals/Goals_sum,
         behinds.ratio=Behinds/Behinds_sum,
         hitouts.ratio=Hit.Outs/Hit.Outs_sum,
         tackles.ratio=Tackles/Tackles_sum,
         rebounds.ratio=Rebounds/Rebounds_sum,
         inside50s.ratio=Inside.50s/Inside.50s_sum,
         clearances.ratio=Clearances/Clearances_sum,
         clangers.ratio=Clangers/Clangers_sum,
         freefors.ratio=Frees.For/Frees.For_sum,
         freesagainst.ratio=Frees.Against/Frees.Against_sum,
         Contested.Possessions.ratio=Contested.Possessions/Contested.Possessions_sum,
         Uncontested.Possessions.ratio=Uncontested.Possessions/Uncontested.Possessions_sum,
         contested.marks.ratio=Contested.Marks/Contested.Marks_sum,
         marksinside50.ratio=Marks.Inside.50/Marks.Inside.50_sum,
         one.percenters.ratio=One.Percenters/One.Percenters_sum,
         bounces.ratio=Bounces/Bounces_sum,
         goal.assists.ratio=Goal.Assists/Goal.Assists_sum,
         disposals.ratio=(Kicks+Handballs)/(Kicks_sum+Handballs_sum))
df<-complete_df_ratio%>%dplyr::select(Date,Game, First.name, Surname, Season, Round.x, Playing.for,-Brownlow.Votes, Brownlow.Votes_sum,everything())
df<-df%>%dplyr::select(-Brownlow.Votes,everything())
df[is.na(df)] <- 0


in.sample  <- subset(df, Season %in% c(2014:2017))
# out.sample <- subset(df, Season == 2014)
in.sample$Brownlow.Votes <- factor(in.sample$Brownlow.Votes)

in.sample<-in.sample%>%filter(Round.x %in% c("1","2","3","4","5","6","7","8",
                                             "9","10","11","12","13","14","15","16","17","18","19","20","21","22","23","24"))


names(in.sample)

in.sample$Player<-paste(in.sample$First.name,in.sample$Surname)

in.sample<-in.sample%>%dplyr::select(Player, Date, Season, Round.x, Playing.for, margin:Brownlow.Votes)
# in.sample<-in.sample[-c(1,2)]

temp1<-scale(in.sample[,8:29])
in.sample[,8:29]<-temp1
#attributes(temp1)
temp1.center<-attr(temp1,"scaled:center")
temp1.scale<-attr(temp1,"scaled:scale")






summary(in.sample)

fm1<-clm(Brownlow.Votes~ handball.ratio +  Marks.ratio +  
           disposals.ratio+  hitouts.ratio+
           freefors.ratio +  freesagainst.ratio +  tackles.ratio +	Goals.ratio +	behinds.ratio +	Contested.Possessions.ratio+
           Uncontested.Possessions.ratio +	clangers.ratio +	contested.marks.ratio +	marksinside50.ratio +
           clearances.ratio +	rebounds.ratio +	inside50s.ratio +	one.percenters.ratio +	bounces.ratio+
           goal.assists.ratio  +margin, 
         data = in.sample)

fm2<- stepAIC(fm1, direction='backward',type=AIC)


names(fitzRoy::player_stats)

df_2017<-fitzRoy::get_footywire_stats(9514:9709)

team_stats_out<-df_2017%>%
  dplyr::select(Date, Player,Season, Round, Team, CP:SC)%>%
  group_by(Date,Season, Round, Team)%>%
  summarise_if(is.numeric,funs(sum=c(sum(.))))

player_stats_out<-df_2017%>%
  dplyr::select(Date, Player,Season, Round, Team, CP:SC)


complete_df_out<-left_join(player_stats_out,team_stats_out, by=c("Date"="Date", "Season"="Season",  "Team"="Team"))



dataset_scores<-fitzRoy::match_results
names(dataset_scores)
dataset_scores1<-dataset_scores%>%dplyr::select (Date, Round, Home.Team, Home.Points,Game)
dataset_scores2<-dplyr::select(dataset_scores, Date, Round, Away.Team, Away.Points,Game)

colnames(dataset_scores1)[3]<-"Team"
colnames(dataset_scores1)[4]<-"Points"
colnames(dataset_scores2)[3]<-"Team"
colnames(dataset_scores2)[4]<-"Points"


df5<-rbind(dataset_scores1,dataset_scores2)
dataset_margins<-df5%>%group_by(Game)%>%
  arrange(Game)%>%
  mutate(margin=c(-diff(Points),diff(Points)))
dataset_margins$Date<-as.Date(dataset_margins$Date)
complete_df_out$Date<-as.Date(complete_df_out$Date)

dataset_margins<-dataset_margins %>%mutate(Team = str_replace(Team, "Brisbane Lions", "Brisbane"))

dataset_margins<-dataset_margins %>%mutate(Team = str_replace(Team, "Footscray", "Western Bulldogs"))


complete_df_out<-left_join(complete_df_out,dataset_margins,by=c("Date"="Date",  "Team"="Team"))

names(complete_df_out)

####create the new ratios
complete_df_ratio_out<-complete_df_out%>%
  mutate(kick.ratio=K/K_sum,
         Marks.ratio=M/M_sum,
         handball.ratio=HB/HB_sum,
         Goals.ratio=G/G_sum,
         behinds.ratio=B/B_sum,
         hitouts.ratio=HO/HO_sum,
         tackles.ratio=T/T_sum,
         rebounds.ratio=R50/R50_sum,
         inside50s.ratio=I50/I50_sum,
         clearances.ratio=(CCL+SCL)/(CCL_sum+SCL_sum),
         clangers.ratio=CL/CL_sum,
         freefors.ratio=FF/FF_sum,
         freesagainst.ratio=FA/FA_sum,
         Contested.Possessions.ratio=CP/CP_sum,
         Uncontested.Possessions.ratio=UP/UP_sum,
         contested.marks.ratio=CM/CM_sum,
         marksinside50.ratio=MI5/MI5_sum,
         one.percenters.ratio=One.Percenters/One.Percenters_sum,
         bounces.ratio=BO/BO_sum,
         goal.assists.ratio=GA/GA_sum,
         disposals.ratio=D/D_sum)




conforming<-complete_df_ratio_out%>%
  dplyr::select(Player, Date, Season, Round.x, Team, margin, 
                kick.ratio:disposals.ratio)

conforming$Brownlow.Votes<-0
out.sample=conforming
out.sample[is.na(out.sample)] <- 0

newdata   <- out.sample[ , -ncol(out.sample)]

newdata[,6:27]<-scale(newdata[,6:27],center=temp1.center,scale=temp1.scale) 

pre.dict    <- predict(fm2,newdata=newdata, type='prob')
pre.dict.m  <- data.frame(matrix(unlist(pre.dict), nrow= nrow(newdata)))
colnames(pre.dict.m) <- c("vote.0", "vote.1", "vote.2", "vote.3")

newdata.pred  <- cbind.data.frame(newdata, pre.dict.m)



#### Step 1: Get expected value on Votes
newdata.pred$expected.votes <- newdata.pred$vote.1 + 2*newdata.pred$vote.2 + 3*newdata.pred$vote.3

####Join back on matchID whoops!

get_match_ID<-fitzRoy::player_stats

xx<-get_match_ID%>%dplyr::select(Date, Player, Match_id)
newdata.pred<-left_join(newdata.pred, xx, by=c("Date"="Date",  "Player"="Player"))



# newdata.pred<-filter(newdata.pred, Date<"2018-09-01")


sum1 <- aggregate(vote.1~Match_id, data = newdata.pred, FUN = sum ); names(sum1) <- c("Match_id", "sum.vote.1");
sum2 <- aggregate(vote.2~Match_id, data = newdata.pred, FUN = sum ); names(sum2) <- c("Match_id", "sum.vote.2");
sum3 <- aggregate(vote.3~Match_id, data = newdata.pred, FUN = sum ); names(sum3) <- c("Match_id", "sum.vote.3");

#### Step 3: Add sum of each vote by matchId to big table
newdata.pred <- merge(newdata.pred, sum1, by = "Match_id")
newdata.pred <- merge(newdata.pred, sum2, by = "Match_id")
newdata.pred <- merge(newdata.pred, sum3, by = "Match_id")

#### Step 4: Add std1/2/3
newdata.pred$std.1  <- (newdata.pred$sum.vote.1/newdata.pred$vote.1)^-1
newdata.pred$std.2  <- (newdata.pred$sum.vote.2/newdata.pred$vote.2)^-1
newdata.pred$std.3  <- (newdata.pred$sum.vote.3/newdata.pred$vote.3)^-1


#### Step 5: Expected standard game vote
newdata.pred$exp_std_game_vote <- newdata.pred$std.1 + 2*newdata.pred$std.2 + 3*newdata.pred$std.3  


#### Step 6: List of winners

newdata.pred$PlayerName<-paste(newdata.pred$Player," ",newdata.pred$Team)
winners.stdgame   <- aggregate(exp_std_game_vote~PlayerName, data = newdata.pred, FUN = sum );
winners.stdgame   <- winners.stdgame[order(-winners.stdgame$exp_std_game_vote), ]
winners.stdgame[1:10, ]

```

## Betting tips 


To evaluate the model or a model first lets build a quick one using the above code for 2017 and evaluate it according to the rough metrics that the [Brownlow Professor](https://www.betfair.com.au/hub/brownlow-medal-predictor-2018/) uses.

```{r}
library(fitzRoy)
library(tidyverse)
df<-fitzRoy::get_afltables_stats(start_date = "1897-01-01", end_date = Sys.Date())

df<-df%>%filter(Season>2010)
names(df)
team_stats<-df%>%
  dplyr::select(Date,Home.team, Away.team, First.name,Surname,Season, Round, Playing.for, Kicks:Goal.Assists)%>%
  group_by(Date, Season, Round, Playing.for)%>%
  summarise_if(is.numeric,funs(sum=c(sum(.))))

player_stats<-df%>%
  dplyr::select(Date, First.name,Surname,Season, Round, Playing.for, Kicks:Goal.Assists)

complete_df<-left_join(player_stats,team_stats, by=c("Date"="Date", "Season"="Season",  "Playing.for"="Playing.for"))

#but we also need margins as per honours stuff

dataset_scores<-fitzRoy::match_results
names(dataset_scores)
dataset_scores1<-dataset_scores%>%dplyr::select (Date, Round, Home.Team, Home.Points,Game)
dataset_scores2<-dplyr::select(dataset_scores, Date, Round, Away.Team, Away.Points,Game)

colnames(dataset_scores1)[3]<-"Team"
colnames(dataset_scores1)[4]<-"Points"
colnames(dataset_scores2)[3]<-"Team"
colnames(dataset_scores2)[4]<-"Points"

df5<-rbind(dataset_scores1,dataset_scores2)
dataset_margins<-df5%>%group_by(Game)%>%
  arrange(Game)%>%
  mutate(margin=c(-diff(Points),diff(Points)))
# View(dataset_margins)
dataset_margins$Date<-as.Date(dataset_margins$Date)
complete_df$Date<-as.Date(complete_df$Date)

complete_df<-left_join(complete_df,dataset_margins,by=c("Date"="Date",  "Playing.for"="Team"))


complete_df_ratio<-complete_df%>%
  mutate(kick.ratio=Kicks/Kicks_sum,
         Marks.ratio=Marks/Marks_sum,
         handball.ratio=Handballs/Handballs_sum,
         Goals.ratio=Goals/Goals_sum,
         behinds.ratio=Behinds/Behinds_sum,
         hitouts.ratio=Hit.Outs/Hit.Outs_sum,
         tackles.ratio=Tackles/Tackles_sum,
         rebounds.ratio=Rebounds/Rebounds_sum,
         inside50s.ratio=Inside.50s/Inside.50s_sum,
         clearances.ratio=Clearances/Clearances_sum,
         clangers.ratio=Clangers/Clangers_sum,
         freefors.ratio=Frees.For/Frees.For_sum,
         freesagainst.ratio=Frees.Against/Frees.Against_sum,
         Contested.Possessions.ratio=Contested.Possessions/Contested.Possessions_sum,
         Uncontested.Possessions.ratio=Uncontested.Possessions/Uncontested.Possessions_sum,
         contested.marks.ratio=Contested.Marks/Contested.Marks_sum,
         marksinside50.ratio=Marks.Inside.50/Marks.Inside.50_sum,
         one.percenters.ratio=One.Percenters/One.Percenters_sum,
         bounces.ratio=Bounces/Bounces_sum,
         goal.assists.ratio=Goal.Assists/Goal.Assists_sum,
         disposals.ratio=(Kicks+Handballs)/(Kicks_sum+Handballs_sum))
df<-complete_df_ratio%>%dplyr::select(Date, First.name, Surname, Season, Round.x, Playing.for,-Brownlow.Votes, Brownlow.Votes_sum,everything())
df<-df%>%dplyr::select(-Brownlow.Votes,everything())
df[is.na(df)] <- 0
in.sample  <- subset(df, Season %in% c(2013:2016))
# out.sample <- subset(df, Season == 2014)
in.sample$Brownlow.Votes <- factor(in.sample$Brownlow.Votes)

in.sample<-in.sample%>%filter(Round.x %in% c("1","2","3","4","5","6","7","8",
                                             "9","10","11","12","13","14","15","16","17","18","19","20","21","22","23","24"))


names(in.sample)

in.sample$Player<-paste(in.sample$First.name,in.sample$Surname)

in.sample<-in.sample%>%dplyr::select(Player, Date, Season, Round.x, Playing.for, Home.team, Away.team, margin:Brownlow.Votes)


temp1<-scale(in.sample[,8:29])
in.sample[,8:29]<-temp1
#attributes(temp1)
temp1.center<-attr(temp1,"scaled:center")
temp1.scale<-attr(temp1,"scaled:scale")





library(ordinal)

fm1<-clm(Brownlow.Votes~ handball.ratio +  Marks.ratio +  
           disposals.ratio+  hitouts.ratio+
           freefors.ratio +  freesagainst.ratio +  tackles.ratio +	Goals.ratio +	behinds.ratio +	Contested.Possessions.ratio+
           Uncontested.Possessions.ratio +	clangers.ratio +	contested.marks.ratio +	marksinside50.ratio +
           clearances.ratio +	rebounds.ratio +	inside50s.ratio +	one.percenters.ratio +	bounces.ratio+
           goal.assists.ratio  +margin, 
         data = in.sample)

library(MASS)

fm2<- stepAIC(fm1, direction='backward',type=AIC)

out.sample  <- subset(df, Season %in% c(2017))
out.sample$Brownlow.Votes <- factor(out.sample$Brownlow.Votes)

out.sample<-out.sample%>%filter(Round.x %in% c("1","2","3","4","5","6","7","8",
                                             "9","10","11","12","13","14","15","16","17","18","19","20","21","22","23","24"))


names(out.sample)

out.sample$Player<-paste(out.sample$First.name,out.sample$Surname)

out.sample<-out.sample%>%dplyr::select(Player, Date, Season, Round.x, Playing.for, Home.team, Away.team, margin:Brownlow.Votes)


out.sample[is.na(out.sample)] <- 0

newdata   <- out.sample[ , -ncol(out.sample)]

newdata[,8:29]<-scale(newdata[,8:29],center=temp1.center,scale=temp1.scale) 

pre.dict    <- predict(fm2,newdata=newdata, type='prob')
pre.dict.m  <- data.frame(matrix(unlist(pre.dict), nrow= nrow(newdata)))
colnames(pre.dict.m) <- c("vote.0", "vote.1", "vote.2", "vote.3")

newdata.pred  <- cbind.data.frame(newdata, pre.dict.m)
newdata.pred$expected.votes <- newdata.pred$vote.1 + 2*newdata.pred$vote.2 + 3*newdata.pred$vote.3
newdata.pred$Match_id<-paste(newdata.pred$Season, newdata.pred$Round.x, newdata.pred$Home.team)
sum1 <- aggregate(vote.1~Match_id, data = newdata.pred, FUN = sum ); names(sum1) <- c("Match_id", "sum.vote.1");
sum2 <- aggregate(vote.2~Match_id, data = newdata.pred, FUN = sum ); names(sum2) <- c("Match_id", "sum.vote.2");
sum3 <- aggregate(vote.3~Match_id, data = newdata.pred, FUN = sum ); names(sum3) <- c("Match_id", "sum.vote.3");

#### Step 3: Add sum of each vote by matchId to big table
newdata.pred <- merge(newdata.pred, sum1, by = "Match_id")
newdata.pred <- merge(newdata.pred, sum2, by = "Match_id")
newdata.pred <- merge(newdata.pred, sum3, by = "Match_id")

#### Step 4: Add std1/2/3
newdata.pred$std.1  <- (newdata.pred$sum.vote.1/newdata.pred$vote.1)^-1
newdata.pred$std.2  <- (newdata.pred$sum.vote.2/newdata.pred$vote.2)^-1
newdata.pred$std.3  <- (newdata.pred$sum.vote.3/newdata.pred$vote.3)^-1


#### Step 5: Expected standard game vote
newdata.pred$exp_std_game_vote <- newdata.pred$std.1 + 2*newdata.pred$std.2 + 3*newdata.pred$std.3  


#### Step 6: List of winners

newdata.pred$PlayerName<-paste(newdata.pred$Player," ",newdata.pred$Playing.for)
winners.stdgame   <- aggregate(exp_std_game_vote~PlayerName, data = newdata.pred, FUN = sum );
winners.stdgame   <- winners.stdgame[order(-winners.stdgame$exp_std_game_vote), ]
winners.stdgame[1:10, ]



```

What you might notice about this is that the order is slightly different to what is presented earlier. But what is the different? 
The out.sample uses the afltables data only no footywire data is used at all. In other words the above script just showed that there are differences between the data on AFLtables and footywire!!!!!

Recall earlier, one of the ways that [Brownlow Professor](https://www.betfair.com.au/hub/brownlow-medal-predictor-2018/)  makes an assessment of his model is he looks at home many times that the player ranked within the top 5 get the 3 votes 95% of the time, 84% of the time for the 2 votes and 74% of the time for the one vote. 

Well using the above method, we could check it for the 2017 AFL season. 

To do this we need to join on the actual votes that happened in 2018 to the predicted probabilities based on our model

```{r}
out.sample  <- subset(df, Season %in% c(2017))
out.sample$Brownlow.Votes <- factor(out.sample$Brownlow.Votes)

out.sample<-out.sample%>%filter(Round.x %in% c("1","2","3","4","5","6","7","8",
                                             "9","10","11","12","13","14","15","16","17","18","19","20","21","22","23","24"))


names(out.sample)

out.sample$Player<-paste(out.sample$First.name,out.sample$Surname)

out.sample<-out.sample%>%dplyr::select(Player, Date, Season, Round.x, Playing.for, Home.team, Away.team, margin:Brownlow.Votes)

fulldataframe<-left_join(newdata.pred,out.sample,by=c("Date"="Date",  "Playing.for"="Playing.for", "Player"="Player"))

```


From there we might want to create a pivot table where we can see how often the highest ranked person by say expected standardised votes gets the 3 votes and so on. 

To do that we can use this bit of script below, it requires a handy package called [janitor](https://cran.r-project.org/web/packages/janitor/index.html)

```{r}


xx<-fulldataframe  %>%
  group_by(Match_id) %>%
  mutate(my_ranks = order(order(exp_std_game_vote, decreasing=TRUE)))
table(xx$my_ranks, xx$Brownlow.Votes)

xx %>% janitor::tabyl(my_ranks, Brownlow.Votes)%>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>% 
  adorn_pct_formatting(rounding = "half up", digits = 2) 
```

So what we can tell from this is over half the time the player we rank first based on our model presented (which I hope you have improved by now) gets the 3 votes. only 32 times out of the 198 games did they fail to poll a single vote. That is not too bad but hopefully your one is better!

## That seems OK but what if I want to see how often a player ranks in x?

That's not too bad from here now that we have gotten most things, we can do that using a filter

```{r}
xx%>%
  dplyr::select(Player, exp_std_game_vote, my_ranks, Brownlow.Votes)%>%
  dplyr::filter(Player=="Dustin Martin")%>%
  janitor::tabyl(my_ranks, Brownlow.Votes)
  

```

What we can see is that out of the 13 times we thought Dustin Martin should poll the most votes he did 10 times, 2 times he polled 1 vote and 1 time he didn't poll. 

These are the kinds of things you might want to look at with your model. For instance, this year everyone is talking up Max Gawn chances. A nice check to do, might be how often do rucks get the 3 votes etc.

Lets show a quick example with Maxy. 

```{r}
df<-fitzRoy::get_afltables_stats(start_date = "1897-01-01", end_date = Sys.Date())
df%>%filter(Season>2010)%>%filter(First.name=="Max", Surname=="Gawn")%>%janitor::tabyl(Brownlow.Votes, Season)
```

Interestingly what we can see is that in Max Gawns other All Aus year (if you believe its a good thing) he polled the 3 votes 4 times. Was this year his better and if so should he poll more?

Lets compare his season in 2016 to his season in 2018

```{r}

library(tidyverse)
library(fitzRoy)
df<-fitzRoy::player_stats
df<-df%>%
  filter(Season != 2018) #filters out the 2018 data (incomeplete that was downloaded when installing fitzRoy for first time) 
df1<-fitzRoy::get_footywire_stats(9514:9709) 
df2<-rbind(df, df1)
df3<-df2%>%
  filter(Season %in% c(2016,2018))%>%
  filter(Player =="Max Gawn") 

df3<-df2%>%
  filter(Season %in% c(2016,2018))%>%
  filter(Player =="Max Gawn") %>%
  group_by(Season)%>%
  summarise(ave.ho=mean(HO),
            ave.CM=mean(CM),
            ave.SC=mean(SC),
            ave.MG=mean(MG),
            ave.ITC=mean(ITC), 
            ave.AF=mean(AF),
            ave.SC=mean(SC),
            ave.Mi5=mean(MI5))

df4<-gather(df3,variables, values, -Season)   
df4%>%   
  ggplot(aes(x=values, y=variables)) +geom_point(aes(colour=as.factor(df4$Season)))+ geom_line(aes(group = variables))
```
This doesn't look very clear because of the different ranges of values, i.e. supercoach scores vs Marks inside 50. 

We should facet this to make it look clearer. 


```{r}
df4%>%   
    ggplot(aes(x=as.factor(df4$Season), y=values)) +
  geom_point()+
  facet_wrap(~variables,scales = "free")


```

So what we can see is that for this group of statistics apart from Marks inside 50, Gawn has done better, does this make him a Brownlow smokey? 

You be the judge. 